---
title: "DataMiningAndWrangling_FA1_KHAFAJI"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

# Lecture 2

## Case Study: Diamonds

```{r diamonds dataset}
diamonds
```


Let’s recall some terminology from Lecture 1:
• Variables: What do variables in this dataset represent? How many of them are there?
• Observations: What do observations in this dataset represent? How many of them are there?
• Values: What are examples of values in this dataset?
• Continuous: Which variables are continuous?
• Categorical: Which variables are categorical?
We can learn more about the variables in this dataset by looking at the documentation:

```{r diamonds documentation}
?diamonds
```

## Exploratory Data Analysis

Before we start modeling our data, it’s a good idea to first explore it. The goals of exploratory data analysis (EDA) are:
• Gaining a basic familiarity with the data
• Exploring the variation of individual variables (the tendency of the values of a variable to change from
observation to observation)
• Exploring the covariation between pairs of variables (the tendency for the values of two or more variables to vary together in a related way)
• Identifying and, if possible, fixing flaws in the data
• Searching for patterns and generating hypotheses
We can explore data by visualizing it or transforming it.

The subject of this lecture is the former: data visualization using ggplot2. The figure below summarizes common strategies for visualizing variation and covariation. We’ll go through each one in this lecture.

## Variation

### Discrete Variables

We usually use bar charts to visualize variation in discrete variables. These can be created through
geom_bar(), which requires the aesthetic x (the variable whose variation we’d like to plot). Let’s take a look at the variation in cut:

```{r Variation-Discrete Variables}

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
         
```
What is the most common kind of diamond cut? 
The answer is the ideal cut.

### Continuous Variables

We usually use histograms to visualize variation in continuous variables. These can be created through geom_histogram(), which requires the aesthetic x (the variable whose variation we’d like to plot). Let’s take a look at the variation in carat:


```{r Variation-Continuous Variables}

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat))
         
```

What can be said about the variation in carat? 
Most diamonds are below 1 carats, but there is a spike in the one carat mark

Most of the values of carat are below 3, so let’s zoom in to that portion of the plot using coord_cartesian:

```{r Variation-Continuous Variables-WithCoordCartesian}

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat)) +
  coord_cartesian(xlim = c(0,3))
         
```

Note the warning message about the number of bins. We might need to experiment to find a meaningful
value for the number of bins in a histogram. Let’s try decreasing the bin width:

```{r Variation-Continuous Variables-WithCoordCartesian-binwidth}

ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.01) +
  coord_cartesian(xlim = c(0,3))
    
```

That’s a peculiar pattern! What kinds of carat values do most diamonds have? Why?
-- Most diamond jewellers tend to make diamonds with carats of whole numbers.


## Covariation

### Discrete vs Discrete

We usually use count plots to visualize covariation between two discrete variables. These represent the number of observations for which the two variables take each combination of values via the size of points. Count plots can be generated using geom_count(), and require x and y aesthetics. For example, let’s assess the covariation between cut and clarity:

```{r Covariation-Discrete-V-Discrete}

ggplot(data = diamonds) +
  geom_count(mapping = aes(x = cut, y = clarity))
    
```
What can we say about the relationship between cut and clarity?
-- Most diamonds have mediocre clarity throughout all the cuts. However, ideal cuts tend to have slightly better clarity.

### Continuous vs Continuous

#### Many to One

We usually use scatter plots to visualize covariation between two continuous variables. These are useful when we can have many y values for each x value. Scatter plots can be generated using geom_point(), and require x and y aesthetics. For example, let’s assess the covariation between x and y:

```{r Covariation-ContVsCont-ManyToOne}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = x, y = y))
```

Whoa! There seem to be some weird outliers in the data. We should probably look into those more and
perhaps remove them. For now, let’s just zoom in on the part of the plot where most of the points lie:

```{r Covariation-ContVsCont-ManyToOne-No_Outliers}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y)) +
  coord_cartesian(xlim = c(3.5, 10), ylim = c(3.5,10))
```

Hm! It looks like most of the points lie near the line y = x. To confirm this, we can add that line to the plot via geom_abline():

```{r Covariation-ContVsCont-ManyToOne-No_Outliers-with ab line}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = x, y = y)) +
geom_abline(slope = 1, intercept = 0, color = "red") +
coord_cartesian(xlim = c(3.5, 10), ylim = c(3.5,10))
```
Why might we have x and y approximately equal for most diamonds in this dataset? 
-- Since x and y are length and width of the diamonds, respectively, most diamond designs aim to have a balanced design, making a square, a diamond shape, or a circle, resulting in similar length and width. 

Why aren’t slope = 1, intercept = 0, and color = "red" inside of an aes()? 
--because we are describing the ab line.

Note that horizontal and vertical lines may be added to plots using geom_hline() and geom_vline(), respectively. Check out the details via ?geom_hline.

#### One to One

```{r Covariation-ContVsCont-one to one-avg price by carat table}
avg_price_by_carat <- diamonds %>%
  mutate(carat = round(carat, 1)) %>%
  group_by(carat) %>%
  summarise(avg_price = mean(price))
avg_price_by_carat
```

To plot avg_price versus carat, we would use a line plot. We can create a line plot via geom_line(),
which requires both x and y aesthetics:

```{r Covariation-Cont vs Cont - One to One}
ggplot(data = avg_price_by_carat) +
geom_line(mapping = aes(x = carat, y = avg_price))
```

What is the relationship between average price and carat? Does this make sense?
--As the carat increases, the average price also increases, although not steadily. It makes sense as a whole, but the drops in prices between 3 carats and 4 carats should be investigates: maybe this is due to an error in the data, or due to variables relating to the sale of the diamonds.


### Continuous vs Discrete

We usually use box plots to visualize covariation between a continuous variable and a discrete variable. A single box plot is a way to visualize the variation in a continuous variable, while side by side box plots allow us to visualize how the distribution of a continuous variable changes based on the values of a discrete variable.

We can construct side by side box plots via geom_boxplot(), which requires x and y aesthetics. For example, let’s see how the price of a diamond depends on its cut:

```{r Covariation- Continuous vs Discrete }
ggplot(data = diamonds) +
geom_boxplot(mapping = aes(x = cut, y = price))
```

Those thick black lines in the middle denote the median price for each cut. What trend do we observe in the median price based as the cut improves? Is this what you would have expected?

-- I am not sure what to think about the huge number of outliers, but given that there are a few diamonds with a high clarity rating, thats to be expected. Since most diamonds are under 1 carat, and have mediocre quality, i think this boxplot gives valuable insight, but is to be expected after analysing the previous graphs.

### Correlation Matrix

To get a more global view of the covariation in a dataset, we might want to visualize its correlation matrix. A correlation is a number between -1 and 1 that captures the strength of association between two variables: 1 means perfectly correlated, 0 means uncorrelated, and -1 means perfectly anticorrelated. A correlation matrix visualizes the correlations between all pairs of variables in a dataset.
Computing the correlation matrix requires a function from the stat471 R package called mixed_cor():

```{r Covariance - Correlation matrix}
library(correlation)
corrmat <- correlation::correlation(diamonds)
corrmat

```
We can plot this correlation matrix using the ggcorrplot() function from the ggcorrplot package:

```{r Covariance - Correlation - Plot}
library(ggcorrplot)
ggcorrplot(corrmat, lab = TRUE, hc.order = TRUE)
```
The ggcorrplot() function is highly customizable. The plot above was created using lab = TRUE to add
the correlation labels and hc.order = TRUE to reorder the variables so that correlated ones appear next to each other.

-- note that the size of the diamond, the carat, and the price are positively correlated.


## Alternative Visualization tools

### Axis Transformations

Some variable are better visualized on a transformed scale, e.g. on a logarithmic scale. A tell-tale sign of a variable that would benefit from a logarithmic transformation is a very long tail, as we saw with carat

```{r Axis Transformation}
ggplot(data = diamonds) +
geom_histogram(aes(x = carat), binwidth = 0.01)
```

We can put the x-axis on a logarithmic scale using scale_x_log10():

```{r Axis Transformation log10}
ggplot(data = diamonds) +
geom_histogram(aes(x = carat), binwidth = 0.01) +
scale_x_log10()
```
The same goes for variables plotted on the y axis. For example:

```{r Axis Transformation log10 x axis and y axis}
ggplot(data = diamonds) +
geom_point(aes(x = carat, y = price)) +
scale_x_log10() +
scale_y_log10()
```
Numerous ways can be made so that we can look at the data at a different perspective.

### Multiple geoms in the same plot

We can add as many geoms as we want to the same plot. We’ve already seen geom_point() and geom_abline() in the same plot. Let’s see another example by adding the individual points to the plot of avg_price versus carat:

```{r multiple geoms in the same plot}
ggplot(data = avg_price_by_carat) +
geom_line(mapping = aes(x = carat, y = avg_price)) +
geom_point(mapping = aes(x = carat, y = avg_price))

```

Note that geom_line() and geom_point() have the same exact aesthetic mapping. Any aesthetic mapping
that applies to all geoms in a plot can be placed inside of ggplot instead of inside the individual geoms. This makes that aesthetic mapping “global”:

```{r global aes}
ggplot(data = avg_price_by_carat, mapping = aes(x = carat, y = avg_price)) +
geom_line() +
geom_point()
```
-- this mapping of different geoms helps us understand our data better, or for this example, the lack of sample size better.

### Multiple aesthetics in the same geom

Any geom can have multiple aesthetics. We’ve already seen this, e.g. specifying both x and y aesthetics for geom_point(). But we can add still more aesthetics, e.g. color, for a richer plot. For example, let’s consider plotting price versus carat:

```{r multiple aesthetics}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = carat, y = price))
```


We can color points based on the cut of the diamond:

```{r multiple aesthetics - color}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = carat, y = price, color = cut))
```
Altering the aesthetics of the geoms help us visualize the data better, making its understanding intuitive.


### Faceting
Each plot so far has contained just one panel. Sometimes, we want to break a plot into multiple panels based on the values of one or two categorical variables in the data. For example, consider again the relationship between price and carat. Instead of coloring points based on cut, we can facet the plot based on cut. When using one faceting variable, we would usually use facet_wrap():

```{r faceting - facet wrap}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = carat, y = price)) +
facet_wrap(~cut)
```

if we want all the panels to be in one row, we can also use facet_grid():

```{r facet grid}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = carat, y = price)) +
facet_grid(. ~ cut)
```

Usually though, facet_grid() is used to facet on two categorical variables, creating a whole matrix of panels. For example:

```{r facet grid multiple}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = carat, y = price)) +
facet_grid(clarity ~ cut)
```
Altering the facets also give us a clear view of the data as it separates the different segmentations.


### Plot Customization

Virtually every aspect of a plot can be customized (plot title, axis titles, legend placement, font sizes, aspect ratio, etc). Work through the Customize Plots tutorial and/or R4DS Chapter 28 to learn more. The most important reason to customize your plots is to make them easier to read and interpret. In addition to correctness, your homework and exams will be graded on presentation quality, including the quality of your plots. See Section 4 of preparing-reports.pdf on Canvas for concrete guidelines.

## References:

ggplot2 cheat sheet
• Visualize Data tutorials
• R4DS Chapters 3 and 28
• preparing reports for STAT 4710


## Exercises

## Relating the carat and cut of a diamond

Recall this plot:

```{r multiple aesthetics - color exercise 7.1, echo=FALSE}
ggplot(data = diamonds) +
geom_point(mapping = aes(x = carat, y = price, color = cut))
```
What relationship does it suggest between carat and cut? Create a plot to directly visualize this relationship. What do you conclude? How does this explain the paradoxical trend we found in the plot below?

-- the plot suggests that the prices rise with the carat, although there is also a lot of carats with ideal cuts and low prices.

```{r faceting - facet wrap - exercise 7.1}
ggplot(data = diamonds) +
geom_boxplot(mapping = aes(x = cut, y = price))
```
the above plot gives a clear picture to the phenomena. Although ideal cuts have the lowest median price among all cuts, premium and very good cuts have the longest whiskers and 75th percentiles.

### Relating the size and carat of a diamond

Create a plot to visualize the relationship between the carat and length of a diamond. Zoom in to exclude any outliers. What sort of relationship does your plot suggest? How would you explain this relationship?
 
```{r exercise 7.2 answer}
ggplot(data = avg_price_by_carat) +
geom_line(mapping = aes(x = carat, y = avg_price)) +
geom_point(mapping = aes(x = carat, y = avg_price))
```

The graph points out that as the carat increases, the price steadily increases, although plateuing at the end.












